First Order Alternation

### 2018-09-25 quantifiers/draft.tex
ok

The problem of checking emptiness of a given automata is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, with no transition quantifiers
\cite{Farzan15}. Since even such simple classes of alternating
automata have no general decision procedure for emptiness, we use an
abstraction-refinement semi-algorithm based on \emph{lazy annotation}
\cite{McMillan06,McMillan14}.

In a nutshell, a lazy annotation procedure systematically explores the
set of execution paths (in our case, sequences of input events) in
search of an accepting execution. Each path has a corresponding path
formula that defines all words accepted along that path. If the
current path formula is satisfiable, the automaton accepts a word and
is not empty. Otherwise, the path is \emph{spurious}. When a spurious
path is encountered, the search backtracks and the path is annotated
with a set of learned facts, that marks this path as spurious in all
future searches. The semi-algorithm uses moreover a coverage relation
between paths, that optimize the search (the continuations of already
covered paths are never explored) and sometimes provide a sound
termination argument, when the automaton is empty.

@@

A lazy annotation semi-algorithm will stop and report emptiness
provided that it succeeds in building a closed and safe unfolding of
the automaton. Notice that, by Definition \ref{def:coverage}, for any
three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
$\beta \cover \gamma$ as well. As we show next (Theorem
\ref{thm:soundness}), there is no need to expand covered nodes,
because, intuitively, there is a word $w \in \lang{\A}$ such that
$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
exists another word $u \in \lang{\A}$ such that $\gamma \preceq
\event{u}$.

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels $U(\beta)$ of (some of the) prefixes $\beta
\prefix \alpha$, including $\alpha$ itself, such that $U(\alpha)$
becomes $\bot$, thus indicating that there is no real execution of the
automaton along that input event sequence. As a result of the change
of labels, if a node $\gamma \prefix \alpha$ used to cover another
node from $\dom(U)$, it might not cover it with the new
label. Therefore, the coverage relation has to be recomputed after
each refinement of the labeling. For a detailed presentation of the
emptiness procedure, we refer the reader to \cite{IosifXu18} or
Appendix \ref{app:pseudocode}.

@@

\begin{theorem}\label{thm:soundness}
  Given an automaton $\A$, if $\A$ has a nonempty safe closed
  unfolding then $\lang{A} = \emptyset$. 
\end{theorem}
\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
  exists a word $w \in \lang{A}$ and let $\alpha \isdef
  \event{w}$.

  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, that there
  exists an interpretation $\I$ such that $\I,\data{w} \models
  \accform{\alpha}$. If $\alpha \in \dom(U)$ then one can show, by
  induction on the length $n\geq0$ of $w$, that $\pathform{\alpha}
  \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w} \models
  \stamp{U(\alpha)}{n}$. But $\I,\data{w} \models \bigwedge_{q \in Q
    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain that
  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
  contradiction.

\qed}

>>>>

The problem of checking emptiness of a given automaton is undecidable,
even for automata with predicates of arity two, whose transition rules
use only equalities and disequalities, having no transition
quantifiers \cite{Farzan15}. Since even such simple classes of
alternating automata have no general decision procedure for emptiness,
we use an abstraction-refinement semi-algorithm based on \emph{lazy
  annotation} \cite{McMillan06,McMillan14}.

In a nutshell, a lazy annotation procedure systematically explores the
set of execution paths (in our case, sequences of input events) in
search of an accepting execution. Each path has a corresponding path
formula that defines all words accepted along that path. If the
current path formula is satisfiable, the automaton accepts a word and
is not empty. Otherwise, the path is said to be \emph{spurious}. When
a spurious path is encountered, the search backtracks and the path is
annotated with a set of learned facts, that marks this path as
spurious in all future searches. The semi-algorithm uses moreover a
coverage relation between paths, that optimize the search (the
continuations of already covered paths are never explored) and
sometimes provide a sound termination argument, when the automaton is
empty.

@@

A lazy annotation semi-algorithm will stop and report emptiness
provided that it succeeds in building a closed and safe unfolding of
the automaton. Notice that, by Definition \ref{def:coverage}, for any
three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
$\beta \cover \gamma$ as well. As we show next (Theorem
\ref{thm:soundness}), there is no need to expand covered nodes,
because, intuitively, there exists a word $w \in \lang{\A}$ such that
$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
exists another word $u \in \lang{\A}$ such that $\gamma \preceq
\event{u}$. Hence, exploring only those input event sequences that are
continuations of $\gamma$ (and ignoring those of $\alpha$) suffices in
order to find a counterexample for emptiness, if one exists.

An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
change (refine) the labels of (some of the) prefixes of $\alpha$ (and
that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
indicating that there is no real execution of the automaton along that
input event sequence. As a result of the change of labels, if a node
$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
might not cover it with the new label. Therefore, the coverage
relation has to be recomputed after each refinement of the
labeling. For a detailed presentation of the emptiness procedure, we
refer the reader to \cite{IosifXu18} (or Appendix
\ref{app:pseudocode}).

@@

\begin{theorem}\label{thm:soundness}
  If an automaton $\A$ has a nonempty safe closed unfolding then
  $\lang{A} = \emptyset$.
\end{theorem}
\proof{ 
  Let $U$ be a safe and complete unfolding of $\A$, such that
  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
  exists a word $w \in \lang{A}$ and let $\alpha \isdef
  \event{w}$.

  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, there
  exists an interpretation $\I$ such that $\I,\data{w} \models
  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
  case, one can show, by induction on the length $n\geq0$ of $w$, that
  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
  contradiction.

  We proceed thus under the assumption that $\alpha \not\in
  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
  $\dom(U)$. Since $U$ is closed, $\alpha'$ is closed and let
  $\alpha_1 \prefix \alpha' \prefix \alpha$ such that $\alpha_1 \cover
  \beta_1$, for some node $\beta_1 \in \dom(U)$. Let $\gamma_1$ be the
  unique sequence such that $\alpha_1\gamma_1 = \alpha$. By Definition
  \ref{def:coverage}, since $\alpha_1 \cover \beta_1$ and $\event{w} =
  \alpha_1\gamma_1 \in \lang{\A}$, there exists a word $w_1$ and a
  cube $c_1 \in \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} =
  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. \qed}
